Characteristics of Top Songs Has Changed from Pandemic Brain

### 2024-04-19 paper/paper.qmd
Updated paper

In the rest of this paper, the data section (@sec-data) will cover the dataset used, how it was obtained, define the variables of interest that are used by our tables and graphs, and briefly explain the data cleaning process. The model section (@sec-model) will explain our proposed causal model, which would explain potential relationships in our data, the setup of our model to understand these relationships, and justification for our model. In the results section (@sec-results), we will reveal tables and graphs made on our datasets, explain what they show, and show our results from our model. In the discussion section (@sec-discussion), we will connect back to the real world and explain what the results could mean, the implications of our results, potential areas of improvement for the paper, and suggestions for future works. Finally, the appendix section (@sec-appendix) will extra tables and graphs from our results as well, as additional information about the model, and a link to a Shiny application featuring interactive graphs of some of the results.

@@

## Methodology
The dataset used in this paper was retrieved, simulated, cleaned, analyzed, and tested using the R programming language [@citeR], tidyverse [@citetidyverse], knitr [@citeknitr], janitor [@citejanitor], dplyr [@citedplyr], ggplot2 [@citeggplot2], spotifyr [@citespotifyr], usethis [@citeusethis], arrow [@citearrow], ggcorrplot [@citeggcorrplot], and testthat [@citetestthat]. Code written in the scripts was checked and styled with lintr [@citelintr] and styler [@citestyler]. Accompanying the paper, a Shiny app [@citeshiny] using shinyWidgets [@citeshinyWidgets] was created containing interactive visualizations. The packages that were used for the model-related sections or used for the model itself are DiagrammeR [@citeDiagrammeR], rsvg [@citersvg], magrittr [@citemagrittr], DiagrammeRsvg [@citeDiagrammeRsvg], png [@citepng], rstanarm [@citerstanarm], and modelsummary [@citemodelsummary].

@@

Each observation in the dataset belongs to a song from the Spotify playlist that we scraped compiled by antoniomendoza and wickeddreamer96 for songs on the Billboard Year-End Hot 100 singles. For the duplicate songs that showed up in different years, they were kept as we wanted to analyze song characteristics. In total, there are 1000 observations in the ```playlists_analysis_data``` dataset, which equals to the number of songs on the Billboard Year-End Hot 100 singles lists from 2014 to 2023.
```{r}
#| label: tbl-dataset-preview
#| tbl-cap: Preview of dataset on Billboard Year-End Hot 100 singles from 2014 to 2023 provided by Spotify
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  select(hit_year, track_duration_ms, tempo, loudness, mode_name, key_mode, period) |>
  head() |>
  kable()
```
Not seen in @tbl-dataset-preview, before_pandemic, major, and minor are numerical versions of the mode_name and period columns. The variables that were originally the raw dataset that was scraped are track_duration_ms, tempo, loudness, mode_name, and key_mode. track_duration_ms describes the duration of the song in milliseconds (ms), tempo is the mean (average) beats per minutes (BPM) of the song, and loudness is mean loudness of a top song in decibels (dB). mode_name represents the key the melody of the song is in, which is either "major" or "minor" and key_mode signifies the scale the song is in.

@@

# Referenced https://www.erikigelstrom.com/articles/causal-graphs-in-r-with-diagrammer/
# https://stackoverflow.com/questions/42737645/how-to-export-images-of-diagrammer-in-r
causal_model <- "digraph {
  graph []
  # Nodes 
  node [shape = plaintext]
    A [label = 'Hit song before 2020 or during and after 2020']
    B [label = 'Modality']
    C [label = 'Loudness']
    D [label = 'Duration']
    E [label = 'Tempo']
    F [label = 'Lyrics (Confounder)']
  # Edges for non-confounding variables
  edge []
    B->A
    C->A
    D->A
    E->A
{ rank = same; }
  # Edges for confounding variable
  edge [style = dashed]
    F -> B
    F -> C
    F -> D
    F -> E
    F -> A
}"

exporting_image <- grViz(causal_model) |>
  export_svg() |>
  charToRaw() |>
  rsvg_png("../other/sketches/graph_2_casual_model.png", width = 1000)

# NOTE: Had to save diagram and read it in as a image since grViz can be only render in HTML and 
# not in PDFs 
# Referenced https://stackoverflow.com/questions/23861000/displaying-images-in-r-in-version-3-1-0
the_image <- readPNG("../other/sketches/graph_2_casual_model.png")
plot.new()
rasterImage(the_image, 0, 0, 0.8, 1.13)
```
A logistic regression model was chosen because our outcome variable, if the song is a top song prior to 2020, is a binary outcome. However, alternative models were considered. Multiple linear regression model was a candidate for our model since the model has multiple predictor variables but it assumes that the outcome is continuous instead of discrete. Other regression models like negative binomial regression were not chosen since we are not dealing with discrete counts for the outcome but instead discrete binary values of 1 and 0. Our logistic regression model describes the following causal relationship (@fig-causal-model) where song characteristics predicts whatever a hit song was made prior to 2020, excluding a confounding variable, lyrics. The assumption we make is that there is a relationship between song characteristics and a top song being a hit prior to 2020. In @fig-causal-model, the reason lyrics, if there are any, is a confounding variable is because lyrics can affect not only if the song was a hit prior to 2020 but also song characteristics from the way the lyrics are performed. However, we are unable to observe the effects of lyrics directly with our data. Weakness and further improvements to the model are discussed in @sec-discussion.

@@

proportion_of_each_mode <- 
  playlist_analysis_data |>
  # Group by if song was a hit before or during and after the pandemic
  group_by(period) |>
  # Count each scale
  count(mode_name) |>
  # rename count column to num_of_key_mode
  rename(
    'num_of_mode_name' = n
  ) |>
  mutate( 
    proportion_mode_name =  round(num_of_mode_name / sum(num_of_mode_name), 2) 
  )

proportion_of_each_mode |> 
  ggplot(mapping = aes(x = mode_name, y = proportion_mode_name)) +
  facet_wrap(facets = vars(period), dir = "h") + 
  geom_bar(stat = "identity", binwidth = 3, fill="#e85569") +
  theme_minimal() +
  labs(x = "Modality",
       y = "Proportion") + 
  theme(axis.text.x = element_text(angle = 90))
```

@@

proportion_of_each_mode |> 
  rename( 
    "Period" = period,
    "Modality" = mode_name,
    "Count of each mode" = num_of_mode_name,
    "Proportion of each mode" = proportion_mode_name
    ) |>
  kable()
```
```{r}
#| label: fig-key-mode-bar
#| fig-cap: "Proportion of songs in different scales before 2020 (the pandemic) versus during and after 2020 (the pandemic) from Billboard's Year-End Hot 100 singles"
#| echo: false
#| warning: false
#| message: false

proportion_of_each_key_mode <- 
  playlist_analysis_data |>
  # Group by if song was a hit before or during and after the pandemic
  group_by(period) |>
  # Count each scale
  count(key_mode) |>
  # rename count column to num_of_key_mode
  rename(
    'num_of_key_mode' = n
  ) |>
  mutate( 
    proportion_key_mode =  round(num_of_key_mode / sum(num_of_key_mode), 2)
  )

proportion_of_each_key_mode |> 
  ggplot(mapping = aes(x = key_mode, y = proportion_key_mode)) +
  facet_wrap(facets = vars(period), dir = "v") + 
  geom_bar(stat="identity", fill="#3E509A") +
  theme_minimal() +
  labs(x = "Scale",
       y = "Proportion") + 
  theme(axis.text.x = element_text(angle = 90))
```
@fig-mode-bar and @tbl-mode-table reveals that a proportion of hit songs before and during and after 2020 that had melodies in a major key was around ~60% (61% before 2020 and 63% from 2020 onwards). Songs in a minor key made up about 40% or less of songs (39% before 2020 and 37% from 2020 onwards) from 2014 to 2023 on Billboard's Year-End Hot 100 singles list. There was a slight increase in songs in a major key from 2020 afterwards however songs in a major key making up a majority of Billboard's Year-End Hot 100 singles stayed consist before 2020 and 2020 onwards.

@@

playlist_analysis_data |> 
  ggplot(aes(x = hit_year, y = track_duration_ms, color = hit_year)) +
  geom_smooth() +
  geom_jitter() +  # So we can see all points
  geom_point(alpha = 0.3) +
  theme_minimal() +
  labs(x = "Hit year",
       y = "Duration of song (in ms)") + 
  geom_vline(xintercept = 2020, linetype = "dashed") +
  scale_color_gradient(low = "#3E509A", high = "#e85569", guide = "none")
```

```{r}
#| label: fig-track-duration-hist
#| fig-cap: "Distribution of track duration (in ms) before 2020 (the pandemic) versus during and after 2020 (the pandemic) from songs on Billboard's Year-End Hot 100 singles list."
#| echo: false
#| warning: false
#| message: false
#| 
playlist_analysis_data |> 
  ggplot(mapping = aes(x = track_duration_ms)) +
  facet_wrap(facets = vars(period), dir = "v") + 
  geom_histogram(binwidth = 7000, fill="#e85569") +
  theme_minimal() +
  labs(x = "Duration of song (in ms)",
       y = "Count")
```
```{r}
#| label: tbl-track-duration-stats-before
#| tbl-cap: "Minimum, quartiles, median, and maximum of track duration (in ms) before 2020 (the pandemic) on Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  # Filter for "Before Pandemic"
  filter(period == "Before Pandemic") |>
  # Choose track_duration_ms
  select(track_duration_ms) |>
  # Rename columns
  rename (
    "Duration of song (in ms)"= track_duration_ms
  ) |>
  summary() |>
  kable()
```

@@

playlist_analysis_data |> 
  # Filter for During and After Pandemic
  filter(period == "During and After Pandemic") |>
  # Choose track_duration_ms
  select(track_duration_ms) |>
  # Rename columns
  rename (
    "Duration of song (in ms)"= track_duration_ms
  ) |>
  summary() |>
  kable()
```
@fig-track-duration-scatterplot reveals that songs over the years from 2014 to 2020 decreased slightly in duration and from 2020 onwards, there was almost no change in trend overall based on the blue, smoother line. @fig-track-duration-hist shows that the distribution for top songs before the pandemic and during and after the pandemic are skewed to the right. According to  @tbl-track-duration-stats-during-after and @tbl-track-duration-stats-before, the distribution of top songs during and after the pandemic are more disperse around their mean of 195999 ms than those from before the pandemic where there is less variance, which means how close data points are to their mean/average value. Based on the distribution, there are more songs with duration of about 216286 ms for top songs before the pandemic. @tbl-track-duration-stats-during-after and @tbl-track-duration-stats-before also show that hit songs during and after the pandemic are shorter in length on average with a mean track duration of 195999 ms compared to hit songs from before the the pandemic with a mean track duration of 216286 ms. The median (the middle value of the data when sorted) track duration also decreased from 214070 ms before 2020 to 195120 ms for 2020 onwards.

### Loudness
```{r}
#| label: fig-loudness-scatterplot
#| fig-cap: "Relationship between the hit year and loudness of a song (in dB) on Billboard's Year-End Hot 100 singles list. Songs from the same year are grouped in the same colour and the horizontal, dashed line divides the hit tracks before 2020 and hit tracks from 2020 and beyond. Blue, smoother line describes the overall trend and relationship between hit year and loudness." 
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  ggplot(aes(x = hit_year, y = loudness, color = hit_year)) +
  geom_smooth() +
  geom_jitter() +  # So we can see all points
  geom_point(alpha = 0.3) +
  theme_minimal() +
  labs(x = "Hit year",
       y = "Loudness of song (in dB)") + 
  geom_vline(xintercept = 2020, linetype = "dashed") +
  scale_color_gradient(low = "#3E509A", high = "#e85569", guide = "none")
```
```{r}
#| label: fig-loudness-hist
#| fig-cap: "Distribution of loudness (in dB) of hit songs before 2020 (the pandemic) versus during and after 2020 (the pandemic) from Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false
playlist_analysis_data |> 
  ggplot(mapping = aes(x = loudness)) +
  facet_wrap(facets = vars(period), dir = "v") + 
  geom_histogram(binwidth = 0.3, fill="#3E509A") +
  theme_minimal() +
  labs(x = "Loudness of song (in dB)",
       y = "Count")
```
```{r}
#| label: tbl-loudness-stats-before
#| tbl-cap: "Minimum, quartiles, median, and maximum of loudness of song (in dB) before 2020 (the pandemic) on Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  # Filter for "Before Pandemic"
  filter(period == "Before Pandemic") |>
  # Choose loudness
  select(loudness) |>
  # Rename columns
  rename (
    "Loudness of song (in dB)"= loudness
  ) |>
  summary() |>
  kable()
```
```{r}
#| label: tbl-loudness-stats-during-after
#| tbl-cap: "Minimum, quartiles, median, and maximum of loudness of song (in dB) during and after 2020 (the pandemic) on Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  # Filter for "During and After Pandemic"
  filter(period == "During and After Pandemic") |>
  # Choose loudness
  select(loudness) |>
  # Rename columns
  rename (
    "Loudness of song (in dB)"= loudness
  ) |>
  summary() |>
  kable()
```
@fig-loudness-scatterplot shows that from 2014 to 2023, there was a decrease in loudness overall with hit songs as see with the negative slope of the blue, smoother line. @fig-loudness-hist reveals that the distribution of top songs' loudness is left-skewed with the loudness of top songs from during and after the pandemic being more disperse than those before the pandemic around their mean. This suggests that top songs from 2020 onwards varied more in terms of their loudness when compared with top songs before 2020. @tbl-loudness-stats-before and @tbl-loudness-stats-during-after shows that before the pandemic, top songs on Billboard's list had a mean loudness of -5.964 dB while during and after the pandemic, it had a mean loudness of -6.516 dB. The median loudness of top tracks has also decrease from -5.700 dB prior to the pandemic to -6.151 dB during and after the pandemic. This indicates that hit songs during and after the pandemic are less louder than hit songs before the pandemic. 
\newpage

### Tempo
```{r}
#| label: fig-tempo-scatterplot
#| fig-cap: "Relationship between the hit year and tempo of a song (the beats per minutes (BPM) of a song) on Billboard's Year-End Hot 100 singles list. Songs from the same year are grouped in the same colour and the horizontal, dashed line divides the hit tracks before 2020 and hit tracks from 2020 and beyond. Blue, smoother line describes the overall trend and relationship between hit year and tempo." 
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  ggplot(aes(x = hit_year, y = tempo, color = hit_year)) +
  geom_smooth() +
  geom_jitter() +  # So we can see all points
  geom_point(alpha = 0.3) +
  theme_minimal() +
  labs(x = "Hit year",
       y = "Tempo of song (BPM)") + 
  geom_vline(xintercept = 2020, linetype = "dashed") +
  scale_color_gradient(low = "#3E509A", high = "#e85569", guide = "none")
```
```{r}
#| label: fig-tempo-hist
#| fig-cap: "Distribution of tempo (BPM) before 2020 (the pandemic) versus during and after 2020 (the pandemic) from Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  ggplot(mapping = aes(x = tempo)) +
  facet_wrap(facets = vars(period), dir = "v") + 
  geom_histogram(binwidth = 3, fill="#e85569") +
  theme_minimal() +
  labs(x = "Tempo of song (BPM)",
       y = "Count")
```
```{r}
#| label: tbl-tempo-stats-before
#| tbl-cap: "Minimum, quartiles, median, and maximum of tempo of a song (BPM) before 2020 (the pandemic) from Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  # Filter for "Before Pandemic"
  filter(period == "Before Pandemic") |>
  # Choose tempo
  select(tempo) |>
  # Rename columns
  rename (
    "Tempo of song (BPM)"= tempo
  ) |>
  summary() |>
  kable()
```
```{r}
#| label: tbl-tempo-stats-during-after
#| tbl-cap: "Minimum, quartiles, median, and maximum of tempo of a song (BPM) during and after 2020 (the pandemic) from Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |> 
  # Filter for "During and After Pandemic"
  filter(period == "During and After Pandemic") |>
  # Choose tempo
  select(tempo) |>
  # Rename columns
  rename (
    "Tempo of song (BPM)"= tempo
  ) |>
  summary() |>
  kable()
```
@fig-tempo-scatterplot reveals almost no change in tempo overall across top songs from 2014 to 2023 based on the blue smoother line appearing to be flat. However, looking at @tbl-tempo-stats-before and @tbl-tempo-stats-during-after it shows that both the median and mean tempo of top songs before 2020 (about 120 BPM and 121 BPM, respectively) are lower or about equal than the median and mean tempo of top songs during and after 2020, which is about 120 BPM and 122 BPM, respectively. @fig-tempo-hist shows that in both periods of time, their distributions have skewness close to zero however there is less dispersion around the mean for hit songs prior to 2020. These results suggest that the the tempo of top songs on Billboard's list are slightly faster during and after 2020.

\newpage
## Model Results

>>>>

In the rest of this paper, the data section (@sec-data) will cover the dataset used, how it was obtained, define the variables of interest that are used by our tables and graphs, and briefly explain the data cleaning process. The model section (@sec-model) will explain our proposed causal model, which would explain potential relationships in our data, the setup of our model to understand these relationships, and justification for our model. In the results section (@sec-results), we will reveal tables and graphs made on our datasets, explain what they show, and show our results from our model. In the discussion section (@sec-discussion), we will explain what we did, connect back to the real world and explain what the results could mean, the implications of our results, potential areas of improvement for the paper, and suggestions for future works. Finally, the appendix section (@sec-appendix) will extra tables and graphs from our results as well, as additional information about the model, and a link to a Shiny application featuring interactive graphs of some of the results.

@@

## Methodology
The dataset used in this paper was retrieved, simulated, cleaned, analyzed, and tested using the R programming language [@citeR], tidyverse [@citetidyverse], knitr [@citeknitr], janitor [@citejanitor], dplyr [@citedplyr], ggplot2 [@citeggplot2], spotifyr [@citespotifyr], usethis [@citeusethis], arrow [@citearrow], ggcorrplot [@citeggcorrplot], and testthat [@citetestthat]. The packages that were used for the model-related sections or used for the model itself are DiagrammeR [@citeDiagrammeR], rsvg [@citersvg], magrittr [@citemagrittr], DiagrammeRsvg [@citeDiagrammeRsvg], png [@citepng], rstanarm [@citerstanarm], and modelsummary [@citemodelsummary].

@@

Each observation in the dataset belongs to a song from the Spotify playlist that we scraped compiled by antoniomendoza and wickeddreamer96 for songs on the Billboard Year-End Hot 100 singles. For the duplicate songs that showed up in different years, they were kept as we wanted to analyze song characteristics. However, we also obtained duplicated songs from the same year-end playlist when calling the Spotify API and so duplicated songs in this case were removed from the dataset. In total, there are 1000 observations in the ```playlists_analysis_data``` dataset, which equals to the number of songs on the Billboard Year-End Hot 100 singles lists from 2014 to 2023.
```{r}
#| label: tbl-dataset-preview
#| tbl-cap: Preview of dataset on Billboard Year-End Hot 100 singles from 2014 to 2023 provided by Spotify
#| echo: false
#| warning: false
#| message: false

# Create table displaying data
playlist_analysis_data |>
  select(hit_year, track_duration_ms, tempo, loudness, mode_name, key_mode, period) |>
  head() |>
  kable()
```
Not seen in @tbl-dataset-preview, before_pandemic, major, and minor are numerical versions of the mode_name and period columns. The variables that were originally the raw dataset that was scraped are track_duration_ms, tempo, loudness, mode_name, and key_mode. track_duration_ms describes the duration of the song in milliseconds (ms), tempo is the mean (average) beats per minutes (BPM) of the song, and loudness is mean loudness of a top song in decibels (dB). mode_name represents the key the melody of the song is in, which is either "major" or "minor" and key_mode signifies the scale the song is in.

@@

# Referenced https://www.erikigelstrom.com/articles/causal-graphs-in-r-with-diagrammer/ and https://stackoverflow.com/questions/42737645/how-to-export-images-of-diagrammer-in-r for code
causal_model <- "digraph {
  graph []
  # Nodes
  node [shape = plaintext]
    A [label = 'Hit song before 2020 or during and after 2020']
    B [label = 'Modality']
    C [label = 'Loudness']
    D [label = 'Duration']
    E [label = 'Tempo']
    F [label = 'Lyrics (Confounder)']
  # Edges for non-confounding variables
  edge []
    B->A
    C->A
    D->A
    E->A
{ rank = same; }
  # Edges for confounding variable
  edge [style = dashed]
    F -> B
    F -> C
    F -> D
    F -> E
    F -> A
}"
# Export the drawing as a png
exporting_image <- grViz(causal_model) |>
  export_svg() |>
  charToRaw() |>
  rsvg_png("../other/sketches/graph_2_casual_model.png", width = 1000)

# NOTE: Had to save diagram and read it in as a image since grViz can be only render in HTML and
# not in PDFs
# Referenced https://stackoverflow.com/questions/23861000/displaying-images-in-r-in-version-3-1-0
# Read in the image and display it
the_image <- readPNG("../other/sketches/graph_2_casual_model.png")
plot.new()
rasterImage(the_image, 0, 0, 0.8, 1.13)
```
A logistic regression model was chosen because our outcome variable, if the song is a top song prior to 2020, is a binary outcome. However, alternative models were considered. Multiple linear regression model was a candidate for our model since the model has multiple predictor variables but it assumes that the outcome is continuous instead of discrete. Other regression models like negative binomial regression were not chosen since we are not dealing with discrete counts for the outcome but instead discrete binary values of 1 and 0. Our logistic regression model describes the following causal relationship (@fig-causal-model) where song characteristics predicts whatever a hit song was made prior to 2020, excluding a confounding variable, lyrics. The assumption we make is that there is a relationship between song characteristics and a top song being a hit prior to 2020. In @fig-causal-model, the reason lyrics, if there are any, is a confounding variable is because lyrics can affect not only if the song was a hit prior to 2020 but also song characteristics from the way the lyrics are performed. However, we are unable to observe the effects of lyrics directly with our data. Weakness and further improvements to the model are discussed in @sec-discussion.

@@

proportion_of_each_mode <-
  playlist_analysis_data |>
  # Group by if song was a hit before or during and after the pandemic
  group_by(period) |>
  # Count each scale
  count(mode_name) |>
  # rename count column to num_of_key_mode
  rename(
    "num_of_mode_name" = n
  ) |>
  mutate(
    proportion_mode_name = round(num_of_mode_name / sum(num_of_mode_name), 2)
  )

# Referenced Telling Stories with Data by Rohan Alexander
# (https://tellingstorieswithdata.com/) for code
proportion_of_each_mode |>
  ggplot(mapping = aes(x = mode_name, y = proportion_mode_name)) +
  # Creates two graphs based on each period
  facet_wrap(facets = vars(period), dir = "h") +
  geom_bar(stat = "identity", binwidth = 3, fill = "#e85569") +
  theme_minimal() +
  labs(
    x = "Modality",
    y = "Proportion"
  ) +
  theme(axis.text.x = element_text(angle = 90))
```

@@

# Create table for dataframe pertaining to modality
proportion_of_each_mode |>
  rename(
    "Period" = period,
    "Modality" = mode_name,
    "Count of each mode" = num_of_mode_name,
    "Proportion of each mode" = proportion_mode_name
  ) |>
  kable()
```
```{r}
#| label: fig-key-mode-bar
#| fig-cap: "Proportion of songs in different scales before 2020 (the pandemic) versus during and after 2020 (the pandemic) from Billboard's Year-End Hot 100 singles"
#| echo: false
#| warning: false
#| message: false

proportion_of_each_key_mode <-
  playlist_analysis_data |>
  # Group by if song was a hit before or during and after the pandemic
  group_by(period) |>
  # Count each scale
  count(key_mode) |>
  # rename count column to num_of_key_mode
  rename(
    "num_of_key_mode" = n
  ) |>
  mutate(
    proportion_key_mode = round(num_of_key_mode / sum(num_of_key_mode), 2)
  )

# Referenced Telling Stories with Data by Rohan Alexander
# (https://tellingstorieswithdata.com/) for code
proportion_of_each_key_mode |>
  ggplot(mapping = aes(x = key_mode, y = proportion_key_mode)) +
  # Creates two graphs based on their period
  facet_wrap(facets = vars(period), dir = "v") +
  geom_bar(stat = "identity", fill = "#3E509A") +
  theme_minimal() +
  labs(
    x = "Scale",
    y = "Proportion"
  ) +
  theme(axis.text.x = element_text(angle = 90))
```
@fig-mode-bar and @tbl-mode-table reveals that a proportion of hit songs before and during and after 2020 that had melodies in a major key was around ~60% (61% before 2020 and 63% from 2020 onwards). Songs in a minor key made up about 40% or less of songs (39% before 2020 and 37% from 2020 onwards) from 2014 to 2023 on Billboard's Year-End Hot 100 singles list. There was a slight increase in songs in a major key from 2020 afterwards however songs in a major key making up a majority of Billboard's Year-End Hot 100 singles stayed consist before 2020 and 2020 onwards.

@@

playlist_analysis_data |>
  ggplot(aes(x = hit_year, y = track_duration_ms, color = hit_year)) +
  geom_smooth() +
  geom_jitter() + # So we can see all points
  geom_point(alpha = 0.3) +
  theme_minimal() +
  labs(
    x = "Hit year",
    y = "Duration of song (in ms)"
  ) +
  geom_vline(xintercept = 2020, linetype = "dashed") +
  scale_color_gradient(low = "#3E509A", high = "#e85569", guide = "none")
```

```{r}
#| label: fig-track-duration-hist
#| fig-cap: "Distribution of track duration (in ms) before 2020 (the pandemic) versus during and after 2020 (the pandemic) from songs on Billboard's Year-End Hot 100 singles list."
#| echo: false
#| warning: false
#| message: false
# Referenced Telling Stories with Data by Rohan Alexander
# (https://tellingstorieswithdata.com/) for code
playlist_analysis_data |>
  ggplot(mapping = aes(x = track_duration_ms)) +
  # Creates two histograms based on period
  facet_wrap(facets = vars(period), dir = "v") +
  geom_histogram(binwidth = 7000, fill = "#e85569") +
  theme_minimal() +
  labs(
    x = "Duration of song (in ms)",
    y = "Count"
  )
```
```{r}
#| label: tbl-track-duration-stats-before
#| tbl-cap: "Minimum, quartiles, median, and maximum of track duration (in ms) before 2020 (the pandemic) on Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |>
  # Filter for "Before Pandemic"
  filter(period == "Before Pandemic") |>
  # Choose track_duration_ms
  select(track_duration_ms) |>
  # Rename columns
  rename(
    "Duration of song (in ms)" = track_duration_ms
  ) |>
  summary() |>
  kable()
```

@@

playlist_analysis_data |>
  # Filter for During and After Pandemic
  filter(period == "During and After Pandemic") |>
  # Choose track_duration_ms
  select(track_duration_ms) |>
  # Rename columns
  rename(
    "Duration of song (in ms)" = track_duration_ms
  ) |>
  summary() |>
  kable()
```
@fig-track-duration-scatterplot reveals that songs over the years from 2014 to 2020 decreased slightly in duration and from 2020 onwards, there was almost no change in trend overall based on the blue, smoother line. @fig-track-duration-hist shows that the distribution for top songs before the pandemic and during and after the pandemic are skewed to the right. According to  @tbl-track-duration-stats-during-after and @tbl-track-duration-stats-before, the distribution of top songs during and after the pandemic are more disperse around their mean of 195999 ms than those from before the pandemic where there is less variance, which means how close data points are to their mean/average value. Based on the distribution, there are more songs with duration of about 216286 ms for top songs before the pandemic. @tbl-track-duration-stats-during-after and @tbl-track-duration-stats-before also show that hit songs during and after the pandemic are shorter in length on average with a mean track duration of 195999 ms compared to hit songs from before the the pandemic with a mean track duration of 216286 ms. The median (the middle value of the data when sorted) track duration also decreased from 214070 ms before 2020 to 195120 ms for 2020 onwards.

### Loudness
```{r}
#| label: fig-loudness-scatterplot
#| fig-cap: "Relationship between the hit year and loudness of a song (in dB) on Billboard's Year-End Hot 100 singles list. Songs from the same year are grouped in the same colour and the horizontal, dashed line divides the hit tracks before 2020 and hit tracks from 2020 and beyond. Blue, smoother line describes the overall trend and relationship between hit year and loudness."
#| echo: false
#| warning: false
#| message: false
# Referenced Telling Stories with Data by Rohan Alexander
# (https://tellingstorieswithdata.com/) for code

# Creates scatterplot
playlist_analysis_data |>
  ggplot(aes(x = hit_year, y = loudness, color = hit_year)) +
  geom_smooth() +
  geom_jitter() + # So we can see all points
  geom_point(alpha = 0.3) +
  theme_minimal() +
  labs(
    x = "Hit year",
    y = "Loudness of song (in dB)"
  ) +
  # Creates vertical line to show divide between top songs before 2020 and 2020
  # onwards
  geom_vline(xintercept = 2020, linetype = "dashed") +
  scale_color_gradient(low = "#3E509A", high = "#e85569", guide = "none")
```
```{r}
#| label: fig-loudness-hist
#| fig-cap: "Distribution of loudness (in dB) of hit songs before 2020 (the pandemic) versus during and after 2020 (the pandemic) from Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false
# Referenced Telling Stories with Data by Rohan Alexander
# (https://tellingstorieswithdata.com/) for code
playlist_analysis_data |>
  ggplot(mapping = aes(x = loudness)) +
  # Creates 2 histograms based on their period
  facet_wrap(facets = vars(period), dir = "v") +
  geom_histogram(binwidth = 0.3, fill = "#3E509A") +
  theme_minimal() +
  labs(
    x = "Loudness of song (in dB)",
    y = "Count"
  )
```
```{r}
#| label: tbl-loudness-stats-before
#| tbl-cap: "Minimum, quartiles, median, and maximum of loudness of song (in dB) before 2020 (the pandemic) on Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |>
  # Filter for "Before Pandemic"
  filter(period == "Before Pandemic") |>
  # Choose loudness
  select(loudness) |>
  # Rename columns
  rename(
    "Loudness of song (in dB)" = loudness
  ) |>
  summary() |>
  kable()
```
```{r}
#| label: tbl-loudness-stats-during-after
#| tbl-cap: "Minimum, quartiles, median, and maximum of loudness of song (in dB) during and after 2020 (the pandemic) on Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |>
  # Filter for "During and After Pandemic"
  filter(period == "During and After Pandemic") |>
  # Choose loudness
  select(loudness) |>
  # Rename columns
  rename(
    "Loudness of song (in dB)" = loudness
  ) |>
  summary() |>
  kable()
```
@fig-loudness-scatterplot shows that from 2014 to 2023, there was a decrease in loudness overall with hit songs as see with the negative slope of the blue, smoother line. @fig-loudness-hist reveals that the distribution of top songs' loudness is left-skewed with the loudness of top songs from during and after the pandemic being more disperse than those before the pandemic around their mean. This suggests that top songs from 2020 onwards varied more in terms of their loudness when compared with top songs before 2020. @tbl-loudness-stats-before and @tbl-loudness-stats-during-after shows that before the pandemic, top songs on Billboard's list had a mean loudness of -5.964 dB while during and after the pandemic, it had a mean loudness of -6.516 dB. The median loudness of top tracks has also decrease from -5.700 dB prior to the pandemic to -6.151 dB during and after the pandemic. This indicates that hit songs during and after the pandemic are less louder than hit songs before the pandemic. 
\newpage

### Tempo
```{r}
#| label: fig-tempo-scatterplot
#| fig-cap: "Relationship between the hit year and tempo of a song (the beats per minutes (BPM) of a song) on Billboard's Year-End Hot 100 singles list. Songs from the same year are grouped in the same colour and the horizontal, dashed line divides the hit tracks before 2020 and hit tracks from 2020 and beyond. Blue, smoother line describes the overall trend and relationship between hit year and tempo."
#| echo: false
#| warning: false
#| message: false
# Referenced Telling Stories with Data by Rohan Alexander
# (https://tellingstorieswithdata.com/) for code

# Creates scatterplot
playlist_analysis_data |>
  ggplot(aes(x = hit_year, y = tempo, color = hit_year)) +
  # Creates smoother line
  geom_smooth() +
  geom_jitter() + # So we can see all points
  geom_point(alpha = 0.3) +
  theme_minimal() +
  labs(
    x = "Hit year",
    y = "Tempo of song (BPM)"
  ) +
  # Creates vertical line to represent the divide between top songs prior to
  # 2020 and 2020 onwards
  geom_vline(xintercept = 2020, linetype = "dashed") +
  scale_color_gradient(low = "#3E509A", high = "#e85569", guide = "none")
```
```{r}
#| label: fig-tempo-hist
#| fig-cap: "Distribution of tempo (BPM) before 2020 (the pandemic) versus during and after 2020 (the pandemic) from Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false
# Referenced Telling Stories with Data by Rohan Alexander
# (https://tellingstorieswithdata.com/) for code
playlist_analysis_data |>
  ggplot(mapping = aes(x = tempo)) +
  # Creates 2 histograms based on their period
  facet_wrap(facets = vars(period), dir = "v") +
  geom_histogram(binwidth = 3, fill = "#e85569") +
  theme_minimal() +
  labs(
    x = "Tempo of song (BPM)",
    y = "Count"
  )
```
```{r}
#| label: tbl-tempo-stats-before
#| tbl-cap: "Minimum, quartiles, median, and maximum of tempo of a song (BPM) before 2020 (the pandemic) from Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |>
  # Filter for "Before Pandemic"
  filter(period == "Before Pandemic") |>
  # Choose tempo
  select(tempo) |>
  # Rename columns
  rename(
    "Tempo of song (BPM)" = tempo
  ) |>
  summary() |>
  kable()
```
```{r}
#| label: tbl-tempo-stats-during-after
#| tbl-cap: "Minimum, quartiles, median, and maximum of tempo of a song (BPM) during and after 2020 (the pandemic) from Billboard's Year-End Hot 100 list"
#| echo: false
#| warning: false
#| message: false

playlist_analysis_data |>
  # Filter for "During and After Pandemic"
  filter(period == "During and After Pandemic") |>
  # Choose tempo
  select(tempo) |>
  # Rename columns
  rename(
    "Tempo of song (BPM)" = tempo
  ) |>
  summary() |>
  kable()
```
@fig-tempo-scatterplot reveals almost no change in tempo overall across top songs from 2014 to 2023 based on the blue smoother line appearing to be flat. However, looking at @tbl-tempo-stats-before and @tbl-tempo-stats-during-after it shows that both the median and mean tempo of top songs before 2020 (about 120 BPM and 121 BPM, respectively) are lower or about equal than the median and mean tempo of top songs during and after 2020, which is about 120 BPM and 122 BPM, respectively. @fig-tempo-hist shows that in both periods of time, their distributions have skewness close to zero however there is less dispersion around the mean for hit songs prior to 2020. These results suggest that the the tempo of top songs on Billboard's list are slightly faster during and after 2020.

\newpage
## Model Results {#sec-model-results}
